---
title: "Identification of differentially expressed isoforms using Ballgown in _S.namaysuch_ non-parasitized/parasitized liver tissue in two different subspecies: lean and siscowet."
author: "Sam White"
date: "08/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Use [Ballgown](https://github.com/alyssafrazee/ballgown) for identification of differentially expressed isoforms in _S.namaysuch_ non-parasitized/parasitized liver tissue in two different subspecies: lean and siscowet.

REQUIRES Linux-based system to run all chunks properly; some chunks will not work on Mac OS!

REQUIRES the following Bash programs:

- `wget`

- `tree`

- `md5sum`

REQUIRES the following R libraries:

- [`Ballgown`](https://github.com/alyssafrazee/ballgown) (Bioconductor)

- `tidyverse`

## Load `R` libraries

```{r}
library("ballgown")
library("tidyverse")
library("ggplot2")
library("Rfast")
```


## Set user variables!
```{r}
# Set maximum pvalue for isoform expression cutoff
pvalue <- 0.05
# Set maximum qvalue (false discovery rate) for isoform expression cutoff
qvalue <- 0.05
```
