---
title: "14-diff-meth"
subtitle: "Differential Methylation Analysis: Siscowet vs Lean Lake Trout"
author: "project-lake-trout"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Overview

This analysis identifies **Differentially Methylated Regions (DMRs)** between 
Siscowet and Lean lake trout (*Salvelinus namaycush*) subspecies using 
CpG methylation data extracted from PacBio HiFi reads.

## Sample Groups

| Group | Samples |
|-------|---------|
| **LEAN** | bc2041, bc2069, bc2070, bc2068 |
| **SISCOWET** | bc2071, bc2073, bc2072, bc2096 |

## Analysis Pipeline

1. Download pre-computed CpG methylation data from [gannet server](https://gannet.fish.washington.edu/v1_web/owlshell/bu-github/project-lake-trout/analyses/10-mCG-call/) (generated by `10-mCG-call.Rmd`)
2. Merge methylation data across all samples
3. Statistical testing (Mann-Whitney U test) for differential methylation
4. FDR correction (Benjamini-Hochberg)
5. Identify Differentially Methylated CpGs (DMCs)
6. Merge nearby DMCs into Differentially Methylated Regions (DMRs)
7. Annotate DMRs with gene information

## Data Source

Methylation data was generated from PacBio HiFi reads using `modkit pileup` and is available at:
<https://gannet.fish.washington.edu/v1_web/owlshell/bu-github/project-lake-trout/analyses/10-mCG-call/>

Files used:
- `*.cov10.bedGraph` - CpG methylation percentages, filtered for coverage ≥ 10

---

# Step 1: Download Methylation Data

Download pre-computed methylation data from the gannet server:

```{bash download-data, eval=FALSE}
# Create directory if it doesn't exist
mkdir -p ../analyses/10-mCG-call

# Download cov10.bedGraph files for all samples
cd ../analyses/10-mCG-call

# Download all sample files
for sample in bc2041 bc2068 bc2069 bc2070 bc2071 bc2072 bc2073 bc2096; do
  if [ ! -f "${sample}.cov10.bedGraph" ]; then
    echo "Downloading ${sample}.cov10.bedGraph..."
    wget -q https://gannet.fish.washington.edu/v1_web/owlshell/bu-github/project-lake-trout/analyses/10-mCG-call/${sample}.cov10.bedGraph
  else
    echo "${sample}.cov10.bedGraph already exists"
  fi
done
```

---

# Step 2: Run Python Analysis Pipeline

The main analysis is performed by the Python script `14-diff-meth.py`.

```{bash run-python, eval=FALSE}
cd ../code
python3 14-diff-meth.py
```

This script will:

- Download methylation data from gannet server (if not already present)
- Perform differential methylation analysis
- Generate output files in `analyses/14-diff-meth/`

---

# Step 3: Load Results for Visualization

```{r load-libraries}
library(tidyverse)
library(ggplot2)
library(knitr)
library(DT)
```

```{r load-results}
# Define paths
results_dir <- "analyses/14-diff-meth"

# Load methylation results if they exist
dmcs_file <- file.path(results_dir, "differential_methylation_cpgs.csv")
dmrs_file <- file.path(results_dir, "dmrs_annotated.csv")
sig_dmcs_file <- file.path(results_dir, "significant_dmcs.csv")

# Check if results exist
results_exist <- file.exists(dmcs_file)

if (results_exist) {
  dmcs <- read_csv(dmcs_file)
  cat("Loaded", nrow(dmcs), "CpG sites\n")
} else {
  cat("Results not found. Please run the Python script first.\n")
}
```

---

# Step 4: Visualize Results

## 4.1 DMC Summary Statistics

```{r dmc-summary, eval=results_exist}
if (exists("dmcs")) {
  # Summary statistics
  summary_stats <- dmcs %>%
    summarise(
      total_cpgs = n(),
      significant_p = sum(significant, na.rm = TRUE),
      significant_fdr = sum(significant_fdr, na.rm = TRUE),
      hyper_siscowet = sum(direction == "hyper_siscowet" & significant, na.rm = TRUE),
      hypo_siscowet = sum(direction == "hypo_siscowet" & significant, na.rm = TRUE)
    )
  
  kable(summary_stats, caption = "Differential Methylation Summary")
}
```

## 4.2 Volcano Plot

```{r volcano-plot, eval=results_exist, fig.width=10, fig.height=8}
if (exists("dmcs")) {
  dmcs_plot <- dmcs %>%
    mutate(
      log10_pval = -log10(pvalue),
      significant_label = case_when(
        significant & direction == "hyper_siscowet" ~ "Hyper in Siscowet",
        significant & direction == "hypo_siscowet" ~ "Hypo in Siscowet",
        TRUE ~ "Not significant"
      )
    )
  
  ggplot(dmcs_plot, aes(x = meth_diff, y = log10_pval, color = significant_label)) +
    geom_point(alpha = 0.5, size = 0.5) +
    scale_color_manual(values = c(
      "Hyper in Siscowet" = "red",
      "Hypo in Siscowet" = "blue",
      "Not significant" = "grey70"
    )) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
    geom_vline(xintercept = c(-15, 15), linetype = "dashed", color = "black") +
    labs(
      title = "Differential Methylation: Siscowet vs Lean",
      x = "Methylation Difference (Siscowet - Lean, %)",
      y = "-log10(p-value)",
      color = "Status"
    ) +
    theme_bw() +
    theme(legend.position = "bottom")
}
```

## 4.3 Methylation Difference Distribution

```{r meth-diff-hist, eval=results_exist, fig.width=10, fig.height=6}
if (exists("dmcs")) {
  sig_dmcs <- dmcs %>% filter(significant)
  
  if (nrow(sig_dmcs) > 0) {
    ggplot(sig_dmcs, aes(x = meth_diff, fill = direction)) +
      geom_histogram(binwidth = 2, alpha = 0.7, position = "identity") +
      scale_fill_manual(values = c("hyper_siscowet" = "red", "hypo_siscowet" = "blue")) +
      labs(
        title = "Distribution of Methylation Differences (Significant DMCs)",
        x = "Methylation Difference (Siscowet - Lean, %)",
        y = "Count",
        fill = "Direction"
      ) +
      theme_bw()
  }
}
```

## 4.4 Chromosome Distribution

```{r chrom-dist, eval=results_exist, fig.width=12, fig.height=6}
if (exists("dmcs")) {
  sig_dmcs <- dmcs %>% filter(significant)
  
  if (nrow(sig_dmcs) > 0) {
    chrom_counts <- sig_dmcs %>%
      group_by(chrom, direction) %>%
      summarise(count = n(), .groups = "drop")
    
    ggplot(chrom_counts, aes(x = reorder(chrom, -count), y = count, fill = direction)) +
      geom_bar(stat = "identity", position = "dodge") +
      scale_fill_manual(values = c("hyper_siscowet" = "red", "hypo_siscowet" = "blue")) +
      labs(
        title = "Significant DMCs per Chromosome",
        x = "Chromosome",
        y = "Number of DMCs",
        fill = "Direction"
      ) +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
  }
}
```

---

# Step 5: DMR Analysis

```{r load-dmrs}
if (file.exists(dmrs_file)) {
  dmrs <- read_csv(dmrs_file)
  cat("Loaded", nrow(dmrs), "DMRs\n")
  dmrs_exist <- TRUE
} else if (file.exists(file.path(results_dir, "dmrs.csv"))) {
  dmrs <- read_csv(file.path(results_dir, "dmrs.csv"))
  cat("Loaded", nrow(dmrs), "DMRs (without gene annotations)\n")
  dmrs_exist <- TRUE
} else {
  cat("DMR results not found.\n")
  dmrs_exist <- FALSE
}
```

## 5.1 DMR Summary

```{r dmr-summary, eval=dmrs_exist}
if (exists("dmrs") && nrow(dmrs) > 0) {
  dmr_summary <- dmrs %>%
    group_by(direction) %>%
    summarise(
      n_dmrs = n(),
      mean_length = round(mean(length), 1),
      mean_cpgs = round(mean(n_cpgs), 1),
      mean_meth_diff = round(mean(abs(mean_diff)), 1)
    )
  
  kable(dmr_summary, caption = "DMR Summary by Direction")
}
```

## 5.2 DMR Size Distribution

```{r dmr-size, eval=dmrs_exist, fig.width=10, fig.height=6}
if (exists("dmrs") && nrow(dmrs) > 0) {
  p1 <- ggplot(dmrs, aes(x = length, fill = direction)) +
    geom_histogram(binwidth = 50, alpha = 0.7, position = "identity") +
    scale_fill_manual(values = c("hyper_siscowet" = "red", "hypo_siscowet" = "blue")) +
    labs(title = "DMR Length Distribution", x = "Length (bp)", y = "Count") +
    theme_bw()
  
  p2 <- ggplot(dmrs, aes(x = n_cpgs, fill = direction)) +
    geom_histogram(binwidth = 1, alpha = 0.7, position = "identity") +
    scale_fill_manual(values = c("hyper_siscowet" = "red", "hypo_siscowet" = "blue")) +
    labs(title = "CpGs per DMR", x = "Number of CpGs", y = "Count") +
    theme_bw()
  
  gridExtra::grid.arrange(p1, p2, ncol = 2)
}
```

## 5.3 Top DMRs Table

```{r top-dmrs, eval=dmrs_exist}
if (exists("dmrs") && nrow(dmrs) > 0) {
  top_dmrs <- dmrs %>%
    arrange(desc(abs(mean_diff))) %>%
    head(20) %>%
    select(dmr_id, chrom, start, end, length, n_cpgs, direction, mean_diff, 
           any_of("overlapping_genes"))
  
  datatable(top_dmrs, 
            caption = "Top 20 DMRs by Methylation Difference",
            options = list(pageLength = 10, scrollX = TRUE))
}
```

---

# Step 6: Gene Overlap Analysis

```{r gene-overlap, eval=dmrs_exist}
if (exists("dmrs") && "overlapping_genes" %in% names(dmrs)) {
  dmrs_with_genes <- dmrs %>%
    filter(overlapping_genes != "" & !is.na(overlapping_genes))
  
  cat("DMRs overlapping genes:", nrow(dmrs_with_genes), "/", nrow(dmrs), "\n")
  
  if (nrow(dmrs_with_genes) > 0) {
    # Extract unique genes
    all_genes <- dmrs_with_genes %>%
      separate_rows(overlapping_genes, sep = ";") %>%
      filter(overlapping_genes != "")
    
    gene_counts <- all_genes %>%
      group_by(overlapping_genes) %>%
      summarise(
        n_dmrs = n(),
        directions = paste(unique(direction), collapse = ","),
        .groups = "drop"
      ) %>%
      arrange(desc(n_dmrs))
    
    datatable(head(gene_counts, 30), 
              caption = "Genes with Most DMR Overlaps")
  }
}
```

---

# Output Files

The analysis generates the following files in `analyses/14-diff-meth/`:

| File | Description |
|------|-------------|
| `methylation_matrix.csv` | Per-sample CpG methylation values |
| `differential_methylation_cpgs.csv` | All tested CpGs with statistics |
| `significant_dmcs.csv` | Significant DMCs only |
| `significant_dmcs.bed` | Significant DMCs in BED format |
| `dmrs.csv` | Differentially methylated regions |
| `dmrs.bed` | DMRs in BED format |
| `dmrs_hyper_siscowet.bed` | Hypermethylated DMRs in Siscowet |
| `dmrs_hypo_siscowet.bed` | Hypomethylated DMRs in Siscowet |
| `dmrs_annotated.csv` | DMRs with gene annotations |
| `methylation_diff_siscowet_vs_lean.bedGraph` | For genome browser |
| `differential_methylation_report.txt` | Summary report |

---

# Session Info

```{r session-info}
sessionInfo()
```

