---
title: "18 - Differential Methylation & PAV: Gene Annotation and Phenotype Synthesis"
subtitle: "Lean vs Siscowet Lake Trout on GCF_016432855.1 (SaNama_1.0)"
author: "project-lake-trout"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r helpers, include=FALSE}
ann <- "analyses/18-annotation"
rd  <- function(f) read.delim(file.path(ann, f), check.names = FALSE,
                              stringsAsFactors = FALSE, quote = "")
kt  <- function(df, cap) knitr::kable(df, row.names = FALSE, caption = cap)
```

# Overview

This report synthesizes the gene-annotation work (Steps 0–4 of
[`code/18-diff-annotation-phenotype-plan.md`](18-diff-annotation-phenotype-plan.md))
into a ranked candidate-gene interpretation and a set of **hypothesized
phenotypic consequences** for the lean and siscowet ecotypes. It does
not generate new variant calls; it asks *where* the existing
differential methylation (DMRs/DMCs) and differential PAVs fall, *what*
those genes do, and *which ecotype phenotypes* they could plausibly
affect.

> ## Read-this-first caveat: the reference is a *lean* genome
>
> GCF_016432855.1 was assembled from a female gynogenetic **doubled
> haploid** of the **Seneca Lake (NY) brood stock** — a documented
> **lean** morphotype (Smith et al. 2022, *Mol Ecol Resour*). Every
> ecotype-asymmetric result here is therefore **lean-reference biased**:
> siscowet diverges more from the reference, inflating apparent
> siscowet-specific PAVs and reducing siscowet CpG coverage in divergent
> regions. Interpret all PAV magnitudes comparatively, never as absolute
> counts, and treat siscowet-specific gene *content* as partially
> invisible to a reference-only analysis.

# Inputs

| Layer | File | Built by |
|------------------------|------------------------|------------------------|
| Gene function table | `gene_function_table.tsv` | `code/18-build-gene-function-table.py` |
| DMR → gene | `dmr_gene_assignments.tsv` | `code/18.1-assign-features-to-genes.py` |
| DMC → gene | `dmc_gene_assignments.tsv` | `code/18.1-...` |
| Stringent PAV → gene | `pav_gene_assignments.tsv` | `code/18.1-...` |
| Lenient PAV burden | `pav_gene_burden.tsv` | `code/18.1-...` |
| Integrated candidates | `integrated_candidate_genes.tsv` | `code/18.2-integrate-candidates.py` |
| GO enrichment | `go_enrichment_{dmr,pav,union}.tsv` | `code/18.3-go-enrichment.py` |

Positional windows: promoter = TSS ± 2 kb (strand-aware); flanking =
gene body ± 5 kb. GO ORA is hypergeometric on local NCBI GO annotations
with DAG propagation, BH-FDR within each set.

# Results

```{r load}
cand <- rd("integrated_candidate_genes.tsv")
dmr  <- rd("dmr_gene_assignments.tsv")
pav  <- rd("pav_gene_assignments.tsv")
go_pav <- rd("go_enrichment_pav.tsv")
num <- function(x) suppressWarnings(as.numeric(x))
```

## Summary

```{r summary}
summ <- data.frame(
  Metric = c("Candidate genes (DMR/DMC/stringent-PAV within 5 kb)",
             "  with a DMR within 5 kb",
             "  with a stringent siscowet-specific deletion within 5 kb",
             "  CONVERGENT (DMR and stringent PAV)",
             "DMRs landing in a promoter (TSS +/- 2 kb)",
             "Genes with an EXON-overlapping siscowet-specific deletion",
             "GO terms FDR<0.1 (PAV set)"),
  Value = c(nrow(cand),
            sum(num(cand$dmr_n) > 0),
            sum(num(cand$pav_strin_n) > 0),
            sum(cand$convergent == 1),
            sum(dmr$in_promoter == 1),
            length(unique(pav$gene_id[pav$exon_overlap == 1])),
            sum(num(go_pav$fdr) < 0.1)))
kt(summ, "Headline numbers")
```

## Convergent genes (methylation *and* structural variation)

The strongest candidates carry both a DMR and a stringent
siscowet-specific deletion.

```{r convergent}
cv <- cand[cand$convergent == 1,
           c("symbol","product","dmr_directions","dmr_best_class","dmr_max_absdiff",
             "pav_strin_best_class","pav_strin_exonic","pav_strin_maxsize","rank_score")]
kt(cv, "Convergent candidate genes")
```

Two of the four (`znf883`-like paralogs, `XlCGF57.1`-like) are
rapidly-evolving, tandemly-arrayed zinc-finger genes — convergence there
partly reflects **repetitive-region mapping**, so they are flagged
rather than asserted.

## Protein-coding genes with promoter methylation

Promoter DMRs are the most interpretable methylation signal (regulatory
potential). Repetitive ncRNA (rDNA/snRNA) loci are excluded.

```{r promoter}
pc <- cand[cand$dmr_promoter == 1 & cand$biotype == "protein_coding" & cand$caution == "",
           c("symbol","product","dmr_directions","dmr_max_absdiff",
             "log2fc_sisco_lean","meth_expr_concordant")]
pc <- pc[order(-num(pc$dmr_max_absdiff)), ]
kt(head(pc, 15), "Protein-coding promoter-DMR genes (top 15 by methylation difference)")
```

## Exonic siscowet-specific deletions (candidate copy-number / loss-of-function)

```{r exonic-pav}
ex <- pav[pav$exon_overlap == 1, c("symbol","product","size_bp","chrom","start","end")]
ex <- ex[order(-num(ex$size_bp)), ]
ex <- ex[!duplicated(ex$symbol), ]
kt(head(ex, 15), "Largest exon-overlapping siscowet-specific deletions (one row per gene)")
```

## GO enrichment — the defensible signal

```{r go}
g <- go_pav[num(go_pav$fdr) < 0.25,
            c("name","namespace","study_k","fold_enrichment","fdr","phenotype")]
g <- g[order(num(g$fdr)), ]
kt(head(g, 20), "PAV-set GO terms at FDR<0.25 (phenotype-relevant flagged in 'phenotype')")
```

Only **calcium ion transport / channel** terms pass FDR\<0.1. Lipid- and
growth-related terms are *suggestive* (FDR 0.1–0.5); broad
`lipid metabolic process` is actually depleted. The calcium signal is
also subject to a **gene-length confounder** (long channel genes
accumulate deletions by chance) — see Limitations.

## Top candidates figure

```{r figure, fig.width=9, fig.height=6, dev='png'}
top <- cand[cand$caution == "" & cand$biotype == "protein_coding", ]
top <- top[order(-num(top$rank_score), -num(top$dmr_max_absdiff)), ]
top <- head(top, 15)
lab <- ifelse(top$symbol == "" | grepl("^LOC", top$symbol),
              substr(top$product, 1, 28), top$symbol)
lab[lab == ""] <- top$gene_id[lab == ""]
col <- ifelse(top$convergent == 1, "#b2182b",
       ifelse(top$pav_strin_exonic == 1, "#ef8a62",
       ifelse(top$dmr_promoter == 1, "#2166ac", "#67a9cf")))
op <- par(mar = c(4, 16, 3, 1))
bp <- barplot(rev(num(top$rank_score)), horiz = TRUE, names.arg = rev(lab),
              las = 1, col = rev(col), xlab = "integrated rank score",
              main = "Top protein-coding candidate genes", cex.names = 0.8)
legend("bottomright", bty = "n", cex = 0.8,
       fill = c("#b2182b","#ef8a62","#2166ac","#67a9cf"),
       legend = c("convergent (meth+PAV)","exonic siscowet deletion",
                  "promoter DMR","other"))
par(op)
dir.create("figures", showWarnings = FALSE)
png("figures/18-top-candidates.png", width = 900, height = 600)
par(mar = c(4, 16, 3, 1))
barplot(rev(num(top$rank_score)), horiz = TRUE, names.arg = rev(lab), las = 1,
        col = rev(col), xlab = "integrated rank score",
        main = "Top protein-coding candidate genes", cex.names = 0.8)
invisible(dev.off())
```

# Phenotype interpretation

Lean and siscowet differ along well-characterized axes: **lipid content
/ energy storage** (siscowet are fatty, deep-water; lean are
lean-bodied, shallow), **body shape and size**, **depth / pressure /
buoyancy**, and **sensory–neuromuscular** performance. The repository's
own `data/measurements.xlsx` is a 17-landmark **geometric-morphometric**
data set (body shape + centroid size) for \~100 fish per group — direct
evidence that the ecotypes differ in *body form and size*, which is the
phenotype the growth/muscle candidates below would feed into. Each block
states the evidence and an explicitly **hypothesized** link.

### Lipid metabolism / energy storage *(gene-level, not genome-wide)*

- **`angptl5`** (angiopoietin-related protein 5-like) — exonic
  siscowet-specific deletion. ANGPTL family members regulate
  **triglyceride / lipoprotein-lipase activity**. Hypothesis: altered
  lipid-handling capacity, consistent with the defining lean/siscowet
  fat-content difference. *Caveat: lean-reference bias means this is a
  candidate, not a confirmed siscowet loss.*
- **`mogat2`** (2-acylglycerol O-acyltransferase 2-A-like) — exonic
  deletion. Catalyses **triacylglycerol synthesis** (monoacylglycerol
  pathway). Hypothesis: differential dietary lipid re-esterification /
  fat deposition.
- **Epoxide hydrolase 1-like** (LOC120029926) — promoter DMR (hypo in
  siscowet). Roles in lipid-epoxide and xenobiotic metabolism.
- **Genome-wide caveat:** GO enrichment does **not** support a broad
  lipid-metabolism program (broad lipid terms are depleted; only narrow
  lipid-binding sub-terms are weakly enriched, FDR\>0.1). The lipid
  signal here is **individual-gene**, not pathway-level.

### Body shape, growth, and muscle

- **`rbm24b`** (RNA-binding motif protein 24b) — promoter DMR,
  **hyper-methylated in siscowet** (Δ33%), the largest-effect promoter
  DMR. Rbm24 is a conserved regulator of **muscle and heart development
  / sarcomere mRNA stability**. Hypothesis: promoter hypermethylation →
  reduced expression → contributes to ecotype differences in musculature
  / body form captured by the morphometric data.
- **Myosin-7-like**, **striated-muscle / contractile-fiber** GO terms,
  and suggestive **developmental-growth** GO enrichment (FDR\~0.15)
  point the same direction.

### Calcium signalling, excitability, and sensory–neural

- **Calcium ion transport** is the one phenotype-relevant GO block
  passing FDR\<0.1 in the PAV set; **calcium-channel** and
  **axon-guidance / neuron-projection** terms top the list. Specific
  genes include a **T-type calcium channel α-1I-like** subunit (exonic
  deletion), **CaMKII β-like** (near a DMR), and **anoctamin-10a**
  (Ca-gated channel, near a DMR). Hypothesis: differences in
  neuromuscular excitability / sensory tuning relevant to deep-water
  (siscowet) vs shallow (lean) life history. *Strong gene-length caveat
  applies.*

### Immune / cell-surface recognition

- Exonic siscowet-specific deletions cluster in **immune & adhesion**
  genes: CEACAM1-like, CD200/OX-2-like, DMBT1-like, NLRP3-like,
  α-2-macroglobulin-like, TRIM25-like, CD22-like. These gene families
  are **rapidly evolving and copy-number-variable** across vertebrates,
  so some of this is expected background rather than ecotype-specific
  selection. Hypothesis: ecotype differences in pathogen/parasite
  exposure (note the companion liver RNAseq is a
  parasitized/non-parasitized study) — but this is the class **most**
  confounded by reference bias and gene-family CNV.

### Chromatin / transcriptional regulation *(flagged as artifact)*

- The DMR GO enrichment (chromatin/nucleosome, transcription-repressor)
  is driven by **one histone gene cluster** (6 co-located histones) and
  **three adjacent `znf883` paralogs** — a tandem-cluster artifact, not
  broad regulatory convergence. Reported for completeness only.

# Limitations

1.  **Reference is lean** — all ecotype-asymmetric signals are
    divergence-biased; siscowet gene content absent from the reference
    is invisible. The per-ecotype hifiasm track (`code/13.3-...plan.md`)
    is the needed complement.
2.  **No DMC survives FDR** (q\<0.1 = 0 in the diff-meth report);
    interpretation leans on DMR-level and stringent-PAV sets. Single-CpG
    and lenient-PAV signals are hypothesis-generating only.
3.  **Expression is orthogonal, not confirmatory** — the liver RNAseq is
    a *different study / different individuals*; most candidates are not
    liver-expressed (`log2fc=NA`).
4.  **Gene-length bias** inflates PAV-based GO enrichment toward long
    genes (calcium channels); a length-matched permutation null is the
    proper test (not yet run).
5.  **Tandem-cluster / paralog inflation** and **IEA-only GO
    annotations** affect enrichment.
6.  **Associative, not causal** — promoter-hypermethylation → silencing
    is a hypothesis; no functional validation, single reference, no
    replication across additional assemblies.

# Conclusions

The cleanest, most defensible findings are: (i) a small set of
**convergent** genes (led by a `znf883` paralog), (ii) a
**calcium-transport / channel** GO signal in the structural-variant set
(with a gene-length caveat), and (iii) **individual** lipid-metabolism
genes (`angptl5`, `mogat2`) and a high-effect **muscle-development**
promoter DMR (`rbm24b`) that align with the known lipid- and body-shape
differences between the ecotypes — the latter directly relevant to the
morphometric phenotypes recorded in `data/measurements.xlsx`. None of
these are confirmed; they are prioritized hypotheses for targeted
follow-up on ecotype-specific (hifiasm) assemblies and in matched
tissue.

```{r session, include=FALSE}
# sessionInfo()
```
