---
title: "20.1-liftoff-annotation"
author: "Steven Roberts"
date: "2026-06-15"
output:
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
citeproc: true
---

# BACKGROUND

Liftoff-only structural annotation of the two ecotype-specific PacBio HiFi assemblies
(`lean`, `siscowet`), per [`code/20-ecotype-genomes.md`](20-ecotype-genomes.md) (Track A).

Because both assemblies are the **same species** as the annotated NCBI reference
**GCF_016432855.1 (SaNama_1.0)**, we transfer the reference annotation directly onto each
assembly with [Liftoff](https://github.com/agshumate/Liftoff) [@shumate2021] rather than
predicting genes de novo. Lifted genes inherit the existing functional annotation in
`analyses/18-annotation/gene_function_table.tsv` via the shared `gene-XXX` ID.

This notebook does **not** do repeat masking or de novo prediction (BRAKER3) — those are
the optional Track B added later.

Due to large file sizes, the outputs are not committed to GitHub. They may be found here:

- [https://gannet.fish.washington.edu/gitrepos/project-lake-trout/output/20.1-liftoff-annotation/](https://gannet.fish.washington.edu/gitrepos/project-lake-trout/output/20.1-liftoff-annotation/)

# SETUP

## Libraries

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  results = "hold",    # Hold outputs and show them after the full code chunk
  warning = FALSE,     # Hide warnings
  collapse = FALSE,    # Keep code and output in separate blocks
  message = FALSE,     # Hide messages
  comment = "##"       # Prefix output lines with '##' so output is visually distinct
)
```

## Set R variables

```{r R-variables, eval=TRUE}
# DIRECTORIES
project_dir <- "/mmfs1/gscratch/scrubbed/samwhite/gitrepos/RobertsLab/project-lake-trout"
genome_dir  <- file.path(project_dir, "data/genome")
ref_dir     <- file.path(project_dir, "data/genome/ref-GCF_016432855.1")
output_dir  <- "../output/20.1-liftoff-annotation"

# INPUT ASSEMBLIES (frozen references for annotation)
lean_fasta     <- file.path(genome_dir, "pb-hifiasm-lean-assembly.fa")
siscowet_fasta <- file.path(genome_dir, "pb-hifiasm-siscowet-assembly.fa")

# REFERENCE ANNOTATION SOURCE
ref_accession  <- "GCF_016432855.1"
ref_fasta      <- file.path(ref_dir, "GCF_016432855.1_SaNama_1.0_genomic.fna")
ref_gff        <- file.path(ref_dir, "genomic.gff")

# EXISTING FUNCTIONAL TABLE (reference genes -> symbol/product/GO)
func_table <- file.path(project_dir, "analyses/18-annotation/gene_function_table.tsv")

# PROGRAMS (Klone /srlab conda envs, matching repo convention)
datasets <- "/srlab/programs/miniforge3-24.7.1-0/envs/ncbi-datasets_env/bin/datasets"
liftoff  <- "/srlab/programs/miniforge3-24.7.1-0/envs/liftoff-1.6.3_env/bin/liftoff"
gffread  <- "/srlab/programs/miniforge3-24.7.1-0/envs/gffread-0.12.7_env/bin/gffread"
samtools <- "/srlab/programs/samtools-1.20/samtools"

# SETTINGS
threads <- "40"

# Export to env vars for bash chunks.
Sys.setenv(
  project_dir    = project_dir,
  genome_dir     = genome_dir,
  ref_dir        = ref_dir,
  output_dir     = output_dir,
  lean_fasta     = lean_fasta,
  siscowet_fasta = siscowet_fasta,
  ref_accession  = ref_accession,
  ref_fasta      = ref_fasta,
  ref_gff        = ref_gff,
  func_table     = func_table,
  datasets       = datasets,
  liftoff        = liftoff,
  gffread        = gffread,
  samtools       = samtools,
  threads        = threads
)
```

```{r make-output-dir, engine='bash'}
mkdir --parents "${output_dir}"
mkdir --parents "${ref_dir}"
```

# REFERENCE ANNOTATION

Download the reference genome FASTA and GFF for `GCF_016432855.1` (SaNama_1.0). This is the
source annotation Liftoff transfers from. Run once; skip if `${ref_gff}` already exists.

```{r download-reference, engine='bash'}
cd "${ref_dir}"

if [ -s "${ref_gff}" ] && [ -s "${ref_fasta}" ]; then
  echo "Reference already present; skipping download."
else
  # NCBI datasets CLI: genome FASTA + GFF3
  "${datasets}" download genome accession "${ref_accession}" \
    --include genome,gff3 \
    --filename "${ref_accession}.zip"

  unzip -o "${ref_accession}.zip"

  # datasets lays files out under ncbi_dataset/data/<accession>/
  cp ncbi_dataset/data/"${ref_accession}"/*_genomic.fna "${ref_fasta}"
  cp ncbi_dataset/data/"${ref_accession}"/genomic.gff   "${ref_gff}"
fi

ls -lh "${ref_dir}"
```

# LIFTOFF

For each ecotype: transfer the reference annotation onto the assembly. Flags:

- `-g` source GFF, `-p` threads
- `-polish` re-aligns to fix gene/exon boundaries (writes `<out>_polished`)
- `-copies -sc 0.95` reports additional gene copies (expansions) at ≥95% identity
- `-u` records reference genes that could **not** be lifted (ecotype-specific loss / PAV candidates)

## Lean

```{r liftoff-lean, engine='bash'}
cd "${output_dir}"

samtools faidx "${lean_fasta}"

"${liftoff}" \
  "${lean_fasta}" \
  "${ref_fasta}" \
  -g "${ref_gff}" \
  -o lean.liftoff.gff3 \
  -u lean.unmapped_features.txt \
  -dir lean_intermediate \
  -p "${threads}" \
  -polish \
  -copies -sc 0.95 \
  > lean.liftoff.log 2>&1

echo "Lean unmapped reference features: $(wc -l < lean.unmapped_features.txt)"
```

## Siscowet

```{r liftoff-siscowet, engine='bash'}
cd "${output_dir}"

samtools faidx "${siscowet_fasta}"

"${liftoff}" \
  "${siscowet_fasta}" \
  "${ref_fasta}" \
  -g "${ref_gff}" \
  -o siscowet.liftoff.gff3 \
  -u siscowet.unmapped_features.txt \
  -dir siscowet_intermediate \
  -p "${threads}" \
  -polish \
  -copies -sc 0.95 \
  > siscowet.liftoff.log 2>&1

echo "Siscowet unmapped reference features: $(wc -l < siscowet.unmapped_features.txt)"
```

# DERIVED FILES

Build the genes-only BED (matching the `data/...genes.bed` convention: `chrom start end
gene-ID 0 strand`) and extract protein FASTAs. `-polish` writes `*.gff3_polished`; use that
when present, otherwise the plain `*.gff3`.

```{r derive-lean, engine='bash'}
cd "${output_dir}"

LEAN_GFF=lean.liftoff.gff3_polished
[ -s "$LEAN_GFF" ] || LEAN_GFF=lean.liftoff.gff3

# Genes-only BED
awk 'BEGIN{OFS="\t"} $3=="gene"{
  id=""; n=split($9,a,";");
  for(i=1;i<=n;i++){ if(a[i]~/^ID=/){ sub(/^ID=/,"",a[i]); id=a[i] } }
  print $1, $4-1, $5, id, 0, $7
}' "$LEAN_GFF" > lean.liftoff.genes.bed

# Protein FASTA
"${gffread}" -y lean.liftoff.proteins.faa -g "${lean_fasta}" "$LEAN_GFF"

echo "Lean lifted genes: $(wc -l < lean.liftoff.genes.bed)"
```

```{r derive-siscowet, engine='bash'}
cd "${output_dir}"

SISCO_GFF=siscowet.liftoff.gff3_polished
[ -s "$SISCO_GFF" ] || SISCO_GFF=siscowet.liftoff.gff3

awk 'BEGIN{OFS="\t"} $3=="gene"{
  id=""; n=split($9,a,";");
  for(i=1;i<=n;i++){ if(a[i]~/^ID=/){ sub(/^ID=/,"",a[i]); id=a[i] } }
  print $1, $4-1, $5, id, 0, $7
}' "$SISCO_GFF" > siscowet.liftoff.genes.bed

"${gffread}" -y siscowet.liftoff.proteins.faa -g "${siscowet_fasta}" "$SISCO_GFF"

echo "Siscowet lifted genes: $(wc -l < siscowet.liftoff.genes.bed)"
```

# INHERIT FUNCTIONAL ANNOTATION

Lifted gene IDs are the reference `gene-XXX` IDs, which are the join key in
`analyses/18-annotation/gene_function_table.tsv`. Join so each ecotype's lifted genes carry
symbol / product / GO with no new computation, and report lift-over coverage.

```{r functional-join, eval=TRUE}
func <- readr::read_tsv(func_table, show_col_types = FALSE)

bed_cols <- c("chrom", "start", "end", "gene_id", "score", "strand")

join_ecotype <- function(bed_path, ecotype) {
  if (!file.exists(bed_path)) {
    message("Skipping ", ecotype, " — ", bed_path, " not found (run Liftoff first).")
    return(invisible(NULL))
  }
  genes <- readr::read_tsv(bed_path, col_names = bed_cols, show_col_types = FALSE)

  out <- genes %>%
    dplyr::select(gene_id, eco_chrom = chrom, eco_start = start,
                  eco_end = end, eco_strand = strand) %>%
    dplyr::left_join(func, by = "gene_id")

  out_path <- file.path(output_dir,
                        paste0(ecotype, ".liftoff.gene_function_table.tsv"))
  readr::write_tsv(out, out_path)

  n_total  <- nrow(out)
  n_func   <- sum(!is.na(out$product) & out$product != "")
  n_named  <- sum(!is.na(out$symbol)  & out$symbol  != "")
  n_go     <- sum(!is.na(out$n_go) & out$n_go > 0)
  ref_n    <- length(unique(func$gene_id))
  lifted_u <- length(unique(out$gene_id))

  message(sprintf(
    "%s: %d gene records (%d unique ref genes, %.1f%% of %d reference genes)\n  %d with product, %d named, %d with GO\n  -> %s",
    ecotype, n_total, lifted_u, 100 * lifted_u / ref_n, ref_n,
    n_func, n_named, n_go, out_path))
}

join_ecotype(file.path(output_dir, "lean.liftoff.genes.bed"), "lean")
join_ecotype(file.path(output_dir, "siscowet.liftoff.genes.bed"), "siscowet")
```

# QC SUMMARY

Quick feature counts and unmapped tallies per ecotype.

```{r qc-summary, engine='bash'}
cd "${output_dir}"

for eco in lean siscowet; do
  GFF=${eco}.liftoff.gff3_polished
  [ -s "$GFF" ] || GFF=${eco}.liftoff.gff3
  echo "=== ${eco} (${GFF}) ==="
  echo "  genes:        $(awk '$3=="gene"' "$GFF" | wc -l)"
  echo "  mRNAs:        $(awk '$3=="mRNA"' "$GFF" | wc -l)"
  echo "  unmapped ref: $(wc -l < ${eco}.unmapped_features.txt)"
done
```

# CHECKSUMS

```{r checksums, engine='bash'}
cd "${output_dir}"

for f in *.gff3 *.gff3_polished *.genes.bed *.proteins.faa *.gene_function_table.tsv; do
  [ -s "$f" ] && md5sum "$f" | tee "$f".md5
done
```

# DELIVERABLES

Per ecotype, written to `output/20.1-liftoff-annotation/`:

| File | Description |
|---|---|
| `{eco}.liftoff.gff3[_polished]` | Lifted gene models on the ecotype assembly |
| `{eco}.liftoff.genes.bed` | Genes-only BED (`data/...genes.bed` convention) |
| `{eco}.liftoff.proteins.faa` | Predicted proteins |
| `{eco}.liftoff.gene_function_table.tsv` | Lifted genes joined to symbol/product/GO |
| `{eco}.unmapped_features.txt` | Reference genes not lifted (PAV / loss candidates) |

These feed the ecotype-coordinate version of
[`code/13.3-hifiasm-differential-methylation-plan.md`](13.3-hifiasm-differential-methylation-plan.md)
and let the PAV results in `code/15-diff-pav.py` be checked against per-ecotype gene models.

# REFERENCES
