---
title: "21.1-gene-anchored-synteny"
author: "Steven Roberts"
date: "2026-06-18"
output:
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
citeproc: true
---

# BACKGROUND

Gene-anchored synteny comparison of the two ecotype assemblies (`lean`, `siscowet`),
following Track B of [`code/21-ecotype-genome-comparison-plan.md`](21-ecotype-genome-comparison-plan.md).

Two things make this pass cheap and clean:

1. **Haplotigs are purged first.** The raw hifiasm assemblies are size-inflated
   (lean 3.73 Gb / 33,035 contigs; siscowet 4.0 Gb / 24,184 contigs vs. ~2.5 Gb expected),
   indicating retained haplotigs that would appear as spurious duplications in any synteny
   map. We run **purge_dups** [@guan2020] on each assembly before annotating, then **re-lift**
   the reference annotation onto the purged sequences (a condensed re-run of
   [`code/20.1-liftoff-annotation.Rmd`](20.1-liftoff-annotation.Rmd)).
2. **Orthology is already solved.** Liftoff assigned every lifted gene the reference
   `gene-XXX` ID. A gene shared by lean and siscowet (or by either ecotype and the reference)
   carries the **same base ID**, so we synthesize MCScanX [@wang2012] anchor pairs directly
   from ID identity — no OrthoFinder / DIAMOND ortholog inference needed. Liftoff `-copies`
   appends a `_N` suffix to extra copies (e.g. `gene-LOC120041454_1`); stripping that suffix
   recovers the base ID used for matching and lets expansions form multiple anchor pairs.

We build three synteny maps: **reference↔lean**, **reference↔siscowet**, and **lean↔siscowet**.
Because both ecotypes anchor to the same 42-chromosome reference, the cross-ecotype comparison
is interpretable despite the fragmented contigs.

Due to large file sizes, outputs are not committed to GitHub. They may be found here:

- [https://gannet.fish.washington.edu/gitrepos/project-lake-trout/output/21.1-gene-anchored-synteny/](https://gannet.fish.washington.edu/gitrepos/project-lake-trout/output/21.1-gene-anchored-synteny/)

# SETUP

## Libraries

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  results = "hold",    # Hold outputs and show them after the full code chunk
  warning = FALSE,     # Hide warnings
  collapse = FALSE,    # Keep code and output in separate blocks
  message = FALSE,     # Hide messages
  comment = "##"       # Prefix output lines with '##' so output is visually distinct
)
```

## Set R variables

```{r R-variables, eval=TRUE}
# DIRECTORIES
project_dir <- "/home/shared/16TB_HDD_01/sr320/github/project-lake-trout"
genome_dir  <- file.path(project_dir, "data/genome")
ref_dir     <- file.path(project_dir, "data/genome/ref-GCF_016432855.1")
reads_dir   <- file.path(project_dir, "data/pacbio-reads/fastq2")
liftoff_dir <- file.path(project_dir, "output/20.1-liftoff-annotation")
output_dir  <- file.path(project_dir, "output/21.1-gene-anchored-synteny")

# RAW INPUT ASSEMBLIES (to be purged)
lean_fasta     <- file.path(genome_dir, "pb-hifiasm-lean-assembly.fa")
siscowet_fasta <- file.path(genome_dir, "pb-hifiasm-siscowet-assembly.fa")

# PacBio HiFi reads per ecotype (barcode -> ecotype per data/pacbio-{lean,siscowet}/README.md)
lean_reads <- paste(
  file.path(reads_dir, "m84082_250614_081210_s3.hifi_reads.demux.bc2041--bc2041.fastq.gz"),
  file.path(reads_dir, "m84082_250614_101514_s4.hifi_reads.demux.bc2069--bc2069.fastq.gz"),
  file.path(reads_dir, "m84082_250614_101514_s4.hifi_reads.demux.bc2070--bc2070.fastq.gz"),
  file.path(reads_dir, "m84082_251001_025627_s2.hifi_reads.bc2068.fastq.gz")
)
siscowet_reads <- paste(
  file.path(reads_dir, "m84082_250614_081210_s3.hifi_reads.demux.bc2071--bc2071.fastq.gz"),
  file.path(reads_dir, "m84082_250614_081210_s3.hifi_reads.demux.bc2072--bc2072.fastq.gz"),
  file.path(reads_dir, "m84082_250614_101514_s4.hifi_reads.demux.bc2073--bc2073.fastq.gz"),
  file.path(reads_dir, "m84082_251001_025627_s2.hifi_reads.bc2096.fastq.gz")
)

# REFERENCE ANNOTATION SOURCE (same as 20.1)
ref_accession <- "GCF_016432855.1"
ref_fasta     <- file.path(ref_dir, "GCF_016432855.1_SaNama_1.0_genomic.fna")
ref_gff       <- file.path(ref_dir, "genomic.gff")
ref_genes_bed <- file.path(project_dir,
                  "data/20220818-snam-GCF_016432855.1_SaNama_1.0_genes.bed")

# PROGRAMS
liftoff      <- "/opt/anaconda/anaconda3/envs/liftoff_env/bin/liftoff"
minimap2     <- "/opt/anaconda/anaconda3/envs/liftoff_env/bin/minimap2"
gffread      <- "/home/shared/gffread-0.12.7.Linux_x86_64/gffread"
samtools     <- "/home/shared/samtools-1.12/samtools"
# Built from source (see Install chunk); adjust to where you put them:
purgedups_bin <- "/home/shared/purge_dups/bin"   # pbcstat, calcuts, split_fa, purge_dups, get_seqs
mcscanx_bin   <- "/home/shared/MCScanX"          # MCScanX, duplicate_gene_classifier

# SETTINGS
threads <- "40"

Sys.setenv(
  project_dir    = project_dir,
  genome_dir     = genome_dir,
  ref_dir        = ref_dir,
  reads_dir      = reads_dir,
  liftoff_dir    = liftoff_dir,
  output_dir     = output_dir,
  lean_fasta     = lean_fasta,
  siscowet_fasta = siscowet_fasta,
  lean_reads     = lean_reads,
  siscowet_reads = siscowet_reads,
  ref_accession  = ref_accession,
  ref_fasta      = ref_fasta,
  ref_gff        = ref_gff,
  ref_genes_bed  = ref_genes_bed,
  liftoff        = liftoff,
  minimap2       = minimap2,
  gffread        = gffread,
  samtools       = samtools,
  purgedups_bin  = purgedups_bin,
  mcscanx_bin    = mcscanx_bin,
  threads        = threads
)
```

```{r make-output-dir, engine='bash'}
mkdir --parents "${output_dir}"
```

## Install tooling (run once)

`purge_dups` and `MCScanX` are not in the existing conda envs. Build them once and point the
`*_bin` variables above at the result. `minimap2`, `liftoff`, `gffread`, `samtools` are already
available (`liftoff_env` + `/home/shared`).

```{r install-tools, engine='bash'}
# purge_dups
cd /home/shared
[ -d purge_dups ] || git clone https://github.com/dfguan/purge_dups.git
cd purge_dups/src && make
ls ../bin   # pbcstat calcuts split_fa purge_dups get_seqs

# MCScanX
cd /home/shared
[ -d MCScanX ] || git clone https://github.com/wyp1125/MCScanX.git
cd MCScanX && make
ls MCScanX duplicate_gene_classifier
```

# STEP 1 — PURGE HAPLOTIGS

Standard purge_dups HiFi pipeline per ecotype: (1) map reads back to the assembly for read-depth,
(2) compute base coverage + automatic cutoffs, (3) self-align the split assembly to find
overlapping haplotigs, (4) purge, (5) extract the purged primary set. `minimap2` takes the four
per-ecotype read files directly (no concatenation needed).

## Lean

```{r purge-lean, engine='bash'}
cd "${output_dir}"
mkdir -p purge/lean && cd purge/lean

# 1. reads -> assembly (HiFi)
"${minimap2}" -t "${threads}" -x map-hifi "${lean_fasta}" ${lean_reads} \
  2> lean.readmap.log | gzip -c > lean.reads.paf.gz

# 2. base coverage + cutoffs
"${purgedups_bin}/pbcstat" lean.reads.paf.gz            # -> PB.base.cov, PB.stat
"${purgedups_bin}/calcuts" PB.stat > cutoffs 2> calcuts.log

# 3. split assembly + self-align
"${purgedups_bin}/split_fa" "${lean_fasta}" > lean.split.fa
"${minimap2}" -t "${threads}" -x asm5 -DP lean.split.fa lean.split.fa \
  2> lean.self.log | gzip -c > lean.split.self.paf.gz

# 4. purge
"${purgedups_bin}/purge_dups" -2 -T cutoffs -c PB.base.cov lean.split.self.paf.gz \
  > lean.dups.bed 2> lean.purge_dups.log

# 5. extract purged primary set (-> purged.fa = primary, hap.fa = removed haplotigs)
"${purgedups_bin}/get_seqs" -e lean.dups.bed "${lean_fasta}" -p lean

mv lean.purged.fa "${output_dir}/lean.purged.fa"
"${samtools}" faidx "${output_dir}/lean.purged.fa"
awk '{s+=$2} END{print "lean purged: "NR" contigs, "s" bp"}' "${output_dir}/lean.purged.fa.fai"
```

## Siscowet

```{r purge-siscowet, engine='bash'}
cd "${output_dir}"
mkdir -p purge/siscowet && cd purge/siscowet

"${minimap2}" -t "${threads}" -x map-hifi "${siscowet_fasta}" ${siscowet_reads} \
  2> siscowet.readmap.log | gzip -c > siscowet.reads.paf.gz

"${purgedups_bin}/pbcstat" siscowet.reads.paf.gz
"${purgedups_bin}/calcuts" PB.stat > cutoffs 2> calcuts.log

"${purgedups_bin}/split_fa" "${siscowet_fasta}" > siscowet.split.fa
"${minimap2}" -t "${threads}" -x asm5 -DP siscowet.split.fa siscowet.split.fa \
  2> siscowet.self.log | gzip -c > siscowet.split.self.paf.gz

"${purgedups_bin}/purge_dups" -2 -T cutoffs -c PB.base.cov siscowet.split.self.paf.gz \
  > siscowet.dups.bed 2> siscowet.purge_dups.log

"${purgedups_bin}/get_seqs" -e siscowet.dups.bed "${siscowet_fasta}" -p siscowet

mv siscowet.purged.fa "${output_dir}/siscowet.purged.fa"
"${samtools}" faidx "${output_dir}/siscowet.purged.fa"
awk '{s+=$2} END{print "siscowet purged: "NR" contigs, "s" bp"}' "${output_dir}/siscowet.purged.fa.fai"
```

# STEP 2 — RE-LIFT ANNOTATION ONTO PURGED ASSEMBLIES

Condensed re-run of [`20.1`](20.1-liftoff-annotation.Rmd) against the purged FASTAs so gene
coordinates match the sequences we compare. Same flags (`-polish -copies -sc 0.95 -u`). Produces
a genes-only BED per ecotype in the `chrom start end gene-ID 0 strand` convention.

```{r relift, engine='bash'}
cd "${output_dir}"

for eco in lean siscowet; do
  ASM="${output_dir}/${eco}.purged.fa"
  "${samtools}" faidx "$ASM"

  "${liftoff}" \
    "$ASM" \
    "${ref_fasta}" \
    -g "${ref_gff}" \
    -o ${eco}.purged.liftoff.gff3 \
    -u ${eco}.purged.unmapped_features.txt \
    -dir ${eco}_purged_intermediate \
    -m "${minimap2}" \
    -p "${threads}" \
    -polish \
    -copies -sc 0.95 \
    > ${eco}.purged.liftoff.log 2>&1

  GFF=${eco}.purged.liftoff.gff3_polished
  [ -s "$GFF" ] || GFF=${eco}.purged.liftoff.gff3

  awk 'BEGIN{OFS="\t"} $3=="gene"{
    id=""; n=split($9,a,";");
    for(i=1;i<=n;i++){ if(a[i]~/^ID=/){ sub(/^ID=/,"",a[i]); id=a[i] } }
    print $1, $4-1, $5, id, 0, $7
  }' "$GFF" > ${eco}.purged.liftoff.genes.bed

  echo "${eco} purged lifted genes: $(wc -l < ${eco}.purged.liftoff.genes.bed)  (unmapped: $(wc -l < ${eco}.purged.unmapped_features.txt))"
done
```

# STEP 3 — BUILD MCScanX INPUTS FROM SHARED IDs

For each pair of genomes, MCScanX needs two files sharing a prefix:

- `<prefix>.gff` — `chrTag<TAB>geneUID<TAB>start<TAB>end`; the first 2 chars of `chrTag` are
  the genome code (`rf` reference, `ln` lean, `si` siscowet), so MCScanX can tell intra- from
  inter-genome blocks.
- `<prefix>.blast` — BLAST `-m8` pairs. **We synthesize these from shared base IDs** instead of
  running an aligner: every gene shared between the two genomes becomes a high-score pair. Extra
  Liftoff copies (`_N`) keep their own UID but share the base ID, so expansions yield multiple
  pairs.

Genome codes and gene UIDs are prefixed (`rf_`, `ln_`, `si_`) to keep names unique in the combined
files. Chromosome/contig labels are stripped of non-alphanumerics (MCScanX dislikes the `.` in
`NC_052307.1`).

```{r mcscanx-inputs, engine='bash'}
cd "${output_dir}"

REF_BED="${ref_genes_bed}"
LEAN_BED="${output_dir}/lean.purged.liftoff.genes.bed"
SISCO_BED="${output_dir}/siscowet.purged.liftoff.genes.bed"

# bed -> MCScanX gff  (tag = 2-char genome code; clean chrom; prefix UID)
bed2gff () {  # $1 bed  $2 code
  awk -v c="$2" 'BEGIN{OFS="\t"}{
    chr=$1; gsub(/[^A-Za-z0-9]/,"",chr);
    print c chr, c"_"$4, $2, $3
  }' "$1"
}
# bed -> "baseID<TAB>UID"  (strip trailing _N copy suffix to get base ID)
bed2map () {  # $1 bed  $2 code
  awk -v c="$2" 'BEGIN{OFS="\t"}{ b=$4; sub(/_[0-9]+$/,"",b); print b, c"_"$4 }' "$1" | sort -k1,1
}

# Precompute per-genome maps
bed2map "$REF_BED"   rf > rf.map
bed2map "$LEAN_BED"  ln > ln.map
bed2map "$SISCO_BED" si > si.map

# pair builder: join two maps on base ID -> reciprocal synthetic blast hits
make_pair () {  # $1 prefix  $2 bedA $3 codeA  $4 bedB $5 codeB  $6 mapA $7 mapB
  bed2gff "$2" "$3"  > "$1.gff"
  bed2gff "$4" "$5" >> "$1.gff"
  join -t"$(printf '\t')" "$6" "$7" \
    | awk 'BEGIN{OFS="\t"}{
        print $2,$3,100,100,0,0,1,100,1,100,"1e-100",500;
        print $3,$2,100,100,0,0,1,100,1,100,"1e-100",500 }' > "$1.blast"
  echo "$1: $(wc -l < $1.blast) synthetic hits, $(wc -l < $1.gff) gene rows"
}

make_pair ref_lean   "$REF_BED" rf "$LEAN_BED"  ln rf.map ln.map
make_pair ref_sisco  "$REF_BED" rf "$SISCO_BED" si rf.map si.map
make_pair lean_sisco "$LEAN_BED" ln "$SISCO_BED" si ln.map si.map
```

# STEP 4 — RUN MCScanX

MCScanX reads `<prefix>.gff` + `<prefix>.blast` (it strips the extensions, so pass the prefix
only) and writes `<prefix>.collinearity` (collinear gene blocks) plus an `html/` directory.
`-a` would emit alignment only; defaults are fine here.

```{r run-mcscanx, engine='bash'}
cd "${output_dir}"

for prefix in ref_lean ref_sisco lean_sisco; do
  echo "=== MCScanX ${prefix} ==="
  "${mcscanx_bin}/MCScanX" "${output_dir}/${prefix}" 2> ${prefix}.mcscanx.log
  echo "  collinear blocks: $(grep -c '^## Alignment' ${prefix}.collinearity)"
  echo "  genes in blocks:  $(grep -cE '^\s*[0-9]+-' ${prefix}.collinearity)"
done
```

# STEP 5 — SUMMARIZE COLLINEARITY

Parse each `.collinearity` into a tidy block table (block id, the two chromosomes/contigs it
spans, number of anchor genes, orientation) and a per-chromosome summary. This is the table that
answers "which ecotype contig maps to which reference chromosome, and is gene order preserved."

```{r summarize, eval=FALSE}
library(tidyverse)

parse_collinearity <- function(path, comparison) {
  lines <- readLines(path)
  blocks <- list(); cur <- NULL
  for (ln in lines) {
    if (str_detect(ln, "^## Alignment")) {
      # e.g. "## Alignment 0: score=... e_value=... N=12 lnptg1&rfNC052307 plus"
      m <- str_match(ln, "Alignment (\\d+):.*N=(\\d+)\\s+(\\S+)&(\\S+)\\s+(plus|minus)")
      cur <- tibble(comparison = comparison, block = as.integer(m[2]),
                    n_anchors = as.integer(m[3]),
                    chr_a = m[4], chr_b = m[5],
                    orientation = m[6])
      blocks[[length(blocks) + 1]] <- cur
    }
  }
  bind_rows(blocks)
}

cmp <- c(ref_lean = "reference_vs_lean",
         ref_sisco = "reference_vs_siscowet",
         lean_sisco = "lean_vs_siscowet")

blocks <- imap_dfr(cmp, ~ parse_collinearity(
  file.path(output_dir, paste0(.y, ".collinearity")), .x))

readr::write_tsv(blocks, file.path(output_dir, "synteny_blocks.tsv"))

# Per-comparison summary
blocks %>%
  group_by(comparison) %>%
  summarise(n_blocks = n(),
            anchor_genes = sum(n_anchors),
            median_block = median(n_anchors),
            inverted_blocks = sum(orientation == "minus")) %>%
  print()

# Contig -> reference chromosome assignment (ecotype-vs-reference comparisons)
blocks %>%
  filter(comparison != "lean_vs_siscowet") %>%
  group_by(comparison, chr_a, chr_b) %>%
  summarise(anchors = sum(n_anchors), .groups = "drop") %>%
  arrange(comparison, desc(anchors)) %>%
  readr::write_tsv(file.path(output_dir, "contig_to_chromosome.tsv"))
```

# STEP 6 — PLOTS

Generate three figure types directly from files already written in this workflow:

1. **Comparison summary bars** from `synteny_blocks.tsv`.
2. **Contig-to-chromosome heatmaps** from `contig_to_chromosome.tsv`.
3. **Gene-level synteny dotplots** from `{ref_lean,ref_sisco,lean_sisco}.collinearity` + `.gff`.

All plots are written to `output/21.1-gene-anchored-synteny/plots/` as PNG and SVG.

```{r plots, eval=TRUE}
library(tidyverse)

plot_dir <- file.path(output_dir, "plots")
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

cmp_meta <- tribble(
  ~prefix,       ~comparison,
  "ref_lean",   "reference_vs_lean",
  "ref_sisco",  "reference_vs_siscowet",
  "lean_sisco", "lean_vs_siscowet"
)

read_gff_mcscan <- function(path) {
  read_tsv(path,
           col_names = c("chr", "gene", "start", "end"),
           col_types = "ccdd",
           show_col_types = FALSE) %>%
    mutate(mid = (start + end) / 2)
}

parse_collinearity_pairs <- function(path, comparison) {
  lines <- readLines(path)
  current <- NULL
  out <- vector("list", length(lines))
  k <- 0

  for (ln in lines) {
    if (str_detect(ln, "^## Alignment")) {
      m <- str_match(
        ln,
        "Alignment (\\d+):.*N=(\\d+)\\s+(\\S+)&(\\S+)\\s+(plus|minus)"
      )
      current <- list(
        block = as.integer(m[, 2]),
        n_anchors = as.integer(m[, 3]),
        chr_a = m[, 4],
        chr_b = m[, 5],
        orientation = m[, 6]
      )
    } else if (str_detect(ln, "^\\s*[0-9]+-\\s*[0-9]+:")) {
      m <- str_match(ln, "^\\s*([0-9]+)-\\s*([0-9]+):\\s+(\\S+)\\s+(\\S+)\\s+([0-9.eE+-]+)$")
      if (!all(is.na(m)) && !is.null(current)) {
        k <- k + 1
        out[[k]] <- tibble(
          comparison = comparison,
          block = current$block,
          n_anchors = current$n_anchors,
          chr_a = current$chr_a,
          chr_b = current$chr_b,
          orientation = current$orientation,
          anchor_idx_a = as.integer(m[, 2]),
          anchor_idx_b = as.integer(m[, 3]),
          gene_a = m[, 4],
          gene_b = m[, 5],
          evalue = m[, 6]
        )
      }
    }
  }

  bind_rows(out[seq_len(k)])
}

# 1) Comparison summary bars
blocks <- read_tsv(file.path(output_dir, "synteny_blocks.tsv"), show_col_types = FALSE)

summary_tbl <- blocks %>%
  group_by(comparison) %>%
  summarise(
    n_blocks = n(),
    anchor_genes = sum(n_anchors),
    pct_inverted = 100 * mean(orientation == "minus"),
    .groups = "drop"
  )

p_summary <- summary_tbl %>%
  pivot_longer(
    cols = c(n_blocks, anchor_genes, pct_inverted),
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = comparison, y = value, fill = comparison)) +
  geom_col(width = 0.75, show.legend = FALSE) +
  facet_wrap(~ metric, scales = "free_y", nrow = 1) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
  labs(
    title = "Synteny block summary",
    x = NULL,
    y = NULL
  )

ggsave(file.path(plot_dir, "01_synteny_summary.png"), p_summary,
       width = 12, height = 3.8, dpi = 300)
ggsave(file.path(plot_dir, "01_synteny_summary.svg"), p_summary,
       width = 12, height = 3.8)

# 2) Contig-to-chromosome heatmap (reference comparisons)
ctc <- read_tsv(file.path(output_dir, "contig_to_chromosome.tsv"), show_col_types = FALSE)

ctc_top <- ctc %>%
  filter(comparison %in% c("reference_vs_lean", "reference_vs_siscowet")) %>%
  group_by(comparison, chr_a) %>%
  summarise(total = sum(anchors), .groups = "drop") %>%
  arrange(comparison, desc(total)) %>%
  group_by(comparison) %>%
  slice_head(n = 60) %>%
  select(comparison, chr_a) %>%
  left_join(ctc, by = c("comparison", "chr_a"))

p_heat <- ctc_top %>%
  ggplot(aes(x = chr_b, y = fct_reorder(chr_a, anchors, .fun = max), fill = anchors)) +
  geom_tile() +
  facet_wrap(~ comparison, scales = "free_y") +
  scale_fill_viridis_c(option = "C", trans = "sqrt") +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Top contig-to-reference chromosome assignments",
    x = "Reference chromosome",
    y = "Ecotype contig",
    fill = "Anchors"
  )

ggsave(file.path(plot_dir, "02_contig_to_chrom_heatmap.png"), p_heat,
       width = 14, height = 11, dpi = 300)
ggsave(file.path(plot_dir, "02_contig_to_chrom_heatmap.svg"), p_heat,
       width = 14, height = 11)

# 3) Gene-level synteny dotplots
for (i in seq_len(nrow(cmp_meta))) {
  prefix <- cmp_meta$prefix[i]
  comparison <- cmp_meta$comparison[i]

  pairs <- parse_collinearity_pairs(
    file.path(output_dir, paste0(prefix, ".collinearity")),
    comparison
  )

  gff <- read_gff_mcscan(file.path(output_dir, paste0(prefix, ".gff")))

  pairs_xy <- pairs %>%
    left_join(gff %>% select(gene, chr, mid), by = c("gene_a" = "gene")) %>%
    rename(chr_a_check = chr, pos_a = mid) %>%
    left_join(gff %>% select(gene, chr, mid), by = c("gene_b" = "gene")) %>%
    rename(chr_b_check = chr, pos_b = mid) %>%
    mutate(
      chr_pair = paste(chr_a, chr_b, sep = " | "),
      orientation = factor(orientation, levels = c("plus", "minus"))
    )

  top_pairs <- pairs_xy %>%
    count(chr_pair, name = "n") %>%
    arrange(desc(n)) %>%
    slice_head(n = 24) %>%
    pull(chr_pair)

  p_dot <- pairs_xy %>%
    filter(chr_pair %in% top_pairs) %>%
    ggplot(aes(x = pos_a / 1e6, y = pos_b / 1e6, color = orientation)) +
    geom_point(size = 0.25, alpha = 0.5) +
    facet_wrap(~ chr_pair, scales = "free", ncol = 6) +
    scale_color_manual(values = c(plus = "#1f77b4", minus = "#d62728")) +
    theme_bw(base_size = 8) +
    theme(
      strip.text = element_text(size = 6),
      legend.position = "top"
    ) +
    labs(
      title = paste("Gene-anchored dotplots:", comparison),
      x = "Genome A gene position (Mb)",
      y = "Genome B gene position (Mb)",
      color = "Orientation"
    )

  ggsave(file.path(plot_dir, paste0("03_dotplot_", comparison, ".png")), p_dot,
         width = 16, height = 10, dpi = 300)
  ggsave(file.path(plot_dir, paste0("03_dotplot_", comparison, ".svg")), p_dot,
         width = 16, height = 10)

  readr::write_tsv(
    pairs_xy,
    file.path(plot_dir, paste0(prefix, ".anchor_pairs.tsv"))
  )
}

list.files(plot_dir)
```

# CHECKSUMS

```{r checksums, engine='bash'}
cd "${output_dir}"
for f in *.purged.fa *.purged.liftoff.genes.bed *.collinearity synteny_blocks.tsv contig_to_chromosome.tsv; do
  [ -s "$f" ] && md5sum "$f" | tee "$f".md5
done
```

# DELIVERABLES

Written to `output/21.1-gene-anchored-synteny/`:

| File | Description |
|---|---|
| `{eco}.purged.fa` | Haplotig-purged primary assembly |
| `{eco}.purged.liftoff.genes.bed` | Re-lifted genes on purged assembly |
| `{ref_lean,ref_sisco,lean_sisco}.collinearity` | MCScanX collinear blocks per comparison |
| `synteny_blocks.tsv` | Tidy block table (chromosomes, anchor counts, orientation) |
| `contig_to_chromosome.tsv` | Ecotype contig → reference chromosome assignment |

These feed the structural-variant / overlay steps in
[`code/21-ecotype-genome-comparison-plan.md`](21-ecotype-genome-comparison-plan.md) (Track A
whole-genome alignment, PAV and DMR overlay).

# REFERENCES
