---
title: "23.1 - Mapping Structural Differences Between Lean and Siscowet Genomes"
format: markdown_github
editor: visual
---

## The question

Lean and siscowet lake trout are two ecotypes of the same species that look and behave
differently — siscowet are deep-water, fat-rich fish; lean are shallower-water and leaner.
Earlier work in this project ([21.1](21.1-gene-anchored-synteny.Rmd)) matched up genes
between the two ecotypes and the reference genome, one gene at a time. That tells us *which*
genes correspond to each other, but not *how the genome itself has been rearranged* —
whether whole chunks of DNA have flipped around, moved to a different chromosome, or been
duplicated. That's what this step, described in full in
[23.1-genome-sv-map.Rmd](23.1-genome-sv-map.Rmd), is for.

## What's a "structural variant"?

Most genetic differences people think of are tiny — a single letter (base) swapped for
another (a SNP). **Structural variants (SVs)** are the bigger, chunkier kind of difference:

- **Inversion** — a segment of DNA is flipped end-to-end.
- **Translocation** — a segment moved to a different location or chromosome.
- **Duplication** — a segment copied, so it appears more than once.
- **Indel** — a stretch of DNA inserted or deleted.
- **Diverged / highly-diverged regions** — stretches that aligned but accumulated a lot of
  small differences.

Any of these can disrupt, duplicate, or relocate genes — which makes them good candidates
for explaining trait differences between ecotypes that a single-gene comparison would miss.

## How we found them

1. **Anchor** each ecotype's genome onto the 42 reference chromosomes using **RagTag**, so
   lean, siscowet, and the reference all share the same chromosome-numbering scheme.
2. **Align** each pair of genomes whole-genome, base-by-base, with **minimap2**.
3. **Call variants** on those alignments with **SyRI**, a tool built specifically to sort
   genome differences into syntenic (matching) blocks versus the SV categories above.
4. **Summarize and plot** the results.

Three comparisons were run: lean-vs-reference, siscowet-vs-reference, and lean-vs-siscowet
directly. (At the time of this writing, siscowet-vs-reference is still processing — the
numbers and plots below cover lean-vs-reference and the direct lean-vs-siscowet comparison.)

## What we found

### How much of the genome is rearranged?

```{r sv-counts, echo=FALSE, out.width="90%", fig.align="center"}
knitr::include_graphics(
  "https://gannet.fish.washington.edu/v1_web/owlshell/bu-github/project-lake-trout/analyses/23-genome-sv/plots/01_sv_counts.png"
)
```

Both comparisons turn up thousands of inversions, duplications, and translocations, plus
hundreds of thousands of small indels — but lean-vs-siscowet (comparing the two ecotypes
directly) consistently shows *more* rearrangements than either ecotype does against the
reference. That's expected: the reference is more closely related to one lineage, so
ecotype-vs-ecotype differences stack the changes from both branches together.

### How much DNA is actually involved?

Counting events is one thing; some SVs are a few bases, others span megabases. This plot
shows the total amount of reference genome sequence (in Mb) covered by each SV class:

```{r sv-span, echo=FALSE, out.width="90%", fig.align="center"}
knitr::include_graphics(
  "https://gannet.fish.washington.edu/v1_web/owlshell/bu-github/project-lake-trout/analyses/23-genome-sv/plots/02_sv_span_mb.png"
)
```

Translocations and highly-diverged regions dominate by footprint — in the lean-vs-siscowet
comparison, translocated segments alone total roughly **230 Mb**, and diverged regions
another **~197 Mb**, out of a ~2.5–2.8 Gb genome. Inversions cover tens of megabases in both
comparisons. So while indels are the most *numerous* SV type, translocations and inversions
are the ones that move around the most sequence.

### What it looks like across the genome

The plots below (from `plotsr`) lay both genomes side by side, chromosome by chromosome, and
draw a ribbon between every matching segment — colored by whether it's syntenic (gray),
inverted, translocated, or duplicated. This is the most intuitive view: you can see directly
which chromosomes have large flipped or relocated blocks.

**Lean vs. reference:**

```{r plotsr-lean-ref, echo=FALSE, out.width="95%", fig.align="center"}
knitr::include_graphics(
  "https://gannet.fish.washington.edu/v1_web/owlshell/bu-github/project-lake-trout/analyses/23-genome-sv/plots/03_plotsr_lean_vs_ref.png"
)
```

**Lean vs. siscowet:**

```{r plotsr-lean-sisco, echo=FALSE, out.width="95%", fig.align="center"}
knitr::include_graphics(
  "https://gannet.fish.washington.edu/v1_web/owlshell/bu-github/project-lake-trout/analyses/23-genome-sv/plots/03_plotsr_lean_vs_sisco.png"
)
```

## Checking our work against an independent method

A separate, earlier analysis ([15-diff-pav.py](15-diff-pav.py)) had already flagged regions
present in one ecotype but absent in another (presence/absence variation, or PAV) just by
comparing coverage against the reference — a much simpler method than SyRI's full alignment.
Cross-validating the two ([23.2-sv-pav-crossvalidate.py](23.2-sv-pav-crossvalidate.py)) asks:
of the regions PAV flagged, how many does SyRI *independently* confirm as a real structural
event?

| Ecotype | PAV-flagged regions | Confirmed by SyRI | Fraction confirmed |
|---|---|---|---|
| lean insertions | 770,891 | 141,159 | 18.3% |
| lean deletions | 225,337 | 45,948 | 20.4% |

Roughly a fifth of the PAV calls line up with a SyRI-confirmed structural variant — mostly
highly-diverged regions, duplications, and translocations. That's a meaningful overlap given
the two methods work completely differently (read-depth/coverage vs. full alignment), and it
gives us a **high-confidence subset** of SVs that two independent approaches agree on.

## Why this matters

This is the first sequence-level, whole-genome map of how the lean and siscowet genomes
differ structurally — not just gene-by-gene, but at the level of entire rearranged segments.
It completes the piece of the ecotype comparison that gene-anchored synteny alone couldn't
answer, and gives us a shortlist of large, high-confidence structural differences to check
against the gene annotations for candidates that might relate to the depth and body-fat
differences between the two ecotypes.

## What's next

- Finish the siscowet-vs-reference comparison and fold it into the summary and plots above.
- Feed the confirmed SVs into the genome browsers alongside the gene-anchored synteny tracks
  ([22-synteny-browser-integration](22-synteny-browser-integration.qmd)), so a rearrangement
  and the genes inside it can be viewed together.
- Look for genes that fall inside the largest, high-confidence inversions/translocations as
  candidates worth a closer look.

All large intermediate files (alignments, full SyRI output, etc.) are too big for GitHub and
are hosted instead at
[gannet.fish.washington.edu/gitrepos/project-lake-trout/analyses/23-genome-sv/](https://gannet.fish.washington.edu/gitrepos/project-lake-trout/analyses/23-genome-sv/).
