---
title: "23.1-genome-sv-map"
author: "Steven Roberts"
date: "2026-07-03"
output:
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
citeproc: true
---

# BACKGROUND

**Phase 1 of [`code/23-next-phase-research-plan.md`](23-next-phase-research-plan.md) — a
whole-genome structural-variant (SV) map between the two ecotypes.** This completes Track A of
[`code/21-ecotype-genome-comparison-plan.md`](21-ecotype-genome-comparison-plan.md), which was
scoped but never run: only the gene-anchored synteny (Track B,
[`21.1`](21.1-gene-anchored-synteny.Rmd)) exists. No inversion, translocation, duplication, or
large-indel calls have been made yet.

The two purged assemblies from [`21.1`](21.1-gene-anchored-synteny.Rmd) are now at biologically
sensible salmonid haploid sizes, so a sequence-level comparison is finally interpretable:

| Assembly | Path | Contigs | Size |
|---|---|---|---|
| lean | `output/21.1-gene-anchored-synteny/lean.purged.fa` | 12,700 | 2.58 Gb |
| siscowet | `output/21.1-gene-anchored-synteny/siscowet.purged.fa` | 8,159 | 2.84 Gb |

**Strategy (identical logic to plan 21).** The purged assemblies are still contig-level (thousands
of contigs), and SyRI [@goel2019] needs near-chromosome-level, homologous, 1-to-1 input. So we
first **anchor** each assembly to the 42 chromosome-scale reference chromosomes
(`GCF_016432855.1`, SaNama_1.0) with **RagTag** [@alonge2022] `scaffold`, giving both ecotypes a
shared 42-chromosome axis where placed scaffolds carry the reference chromosome name. We then run
three pairwise comparisons through that frame — **lean↔reference**, **siscowet↔reference**, and
**lean↔siscowet** — calling SVs with SyRI on whole-genome **minimap2** [@li2018] `asm5`
alignments.

**Why minimap2 and not the MUMmer `nucmer` named in the plan.** The `nucmer` at `/usr/bin` is
MUMmer **v3.1** — single-threaded and impractical on 2.5 Gb genomes. SyRI's own recommended
large-genome input is a `minimap2 -ax asm5 --eqx` alignment, and minimap2 is already the aligner
used throughout this project ([`20.1`](20.1-liftoff-annotation.Rmd),
[`21.1`](21.1-gene-anchored-synteny.Rmd)). We therefore install a modern **MUMmer4** `nucmer`
inside the analysis env for the optional `dnadiff` identity report (Step 6), but drive SV calling
from minimap2.

The cross-validation against the reference-based PAV calls ([`15`](15-diff-pav.py)) is a separate,
lighter step in [`code/23.2-sv-pav-crossvalidate.py`](23.2-sv-pav-crossvalidate.py); this notebook
produces the SV calls it consumes.

Due to large file sizes, outputs are not committed to GitHub. They may be found here:

- [https://gannet.fish.washington.edu/gitrepos/project-lake-trout/analyses/23-genome-sv/](https://gannet.fish.washington.edu/gitrepos/project-lake-trout/analyses/23-genome-sv/)

# SETUP

## Libraries

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  results = "hold",    # Hold outputs and show them after the full code chunk
  warning = FALSE,     # Hide warnings
  collapse = FALSE,    # Keep code and output in separate blocks
  message = FALSE,     # Hide messages
  comment = "##"       # Prefix output lines with '##' so output is visually distinct
)
```

## Set R variables

```{r R-variables, eval=TRUE}
# DIRECTORIES
project_dir <- "/home/shared/16TB_HDD_01/sr320/github/project-lake-trout"
ref_dir     <- file.path(project_dir, "data/genome/ref-GCF_016432855.1")
synteny_dir <- file.path(project_dir, "output/21.1-gene-anchored-synteny")
output_dir  <- file.path(project_dir, "analyses/23-genome-sv")

# PURGED ECOTYPE ASSEMBLIES (from 21.1)
lean_fasta     <- file.path(synteny_dir, "lean.purged.fa")
siscowet_fasta <- file.path(synteny_dir, "siscowet.purged.fa")

# CHROMOSOME-SCALE REFERENCE (the anchor)
ref_fasta <- file.path(ref_dir, "GCF_016432855.1_SaNama_1.0_genomic.fna")

# PROGRAMS
# All tools for this phase live in one env (built below): minimap2, samtools, ragtag, syri,
# plotsr, MUMmer4. Kept as one env because ragtag.py / syri shell out to sibling scripts
# (ragtag_scaffold.py, etc.) and MUMmer wrapper scripts (nucmer, dnadiff) by bare name via
# PATH, not by full path -- so this env's bin/ must be on PATH, not just referenced directly.
syri_bin  <- "/home/sr320/miniconda3/envs/syri_env/bin"
minimap2  <- file.path(syri_bin, "minimap2")
samtools  <- file.path(syri_bin, "samtools")
ragtag    <- file.path(syri_bin, "ragtag.py")
syri      <- file.path(syri_bin, "syri")
plotsr    <- file.path(syri_bin, "plotsr")
dnadiff   <- file.path(syri_bin, "dnadiff")   # MUMmer4

# SETTINGS
threads <- "40"
min_chrom_bp <- "10000000"   # reference sequences > 10 Mb are the 42 chromosomes

# Prepend syri_bin to PATH for this R session so every engine='bash' chunk below (which
# inherits the R process environment) can resolve ragtag's/syri's/dnadiff's sibling scripts.
Sys.setenv(PATH = paste(syri_bin, Sys.getenv("PATH"), sep = .Platform$path.sep))

Sys.setenv(
  project_dir    = project_dir,
  ref_dir        = ref_dir,
  synteny_dir    = synteny_dir,
  output_dir     = output_dir,
  lean_fasta     = lean_fasta,
  siscowet_fasta = siscowet_fasta,
  ref_fasta      = ref_fasta,
  syri_bin       = syri_bin,
  minimap2       = minimap2,
  samtools       = samtools,
  ragtag         = ragtag,
  syri           = syri,
  plotsr         = plotsr,
  dnadiff        = dnadiff,
  threads        = threads,
  min_chrom_bp   = min_chrom_bp
)
```

```{r make-output-dirs, engine='bash'}
mkdir --parents "${output_dir}"/{anchor,wga,syri,plots,logs}
ls -d "${output_dir}"/*
```

## Install tooling (run once)

`ragtag`, `syri`, `plotsr`, and a threaded MUMmer4 `nucmer`/`dnadiff` are not in the existing
conda envs. Create one env for the whole phase. `minimap2` and `samtools` are pulled into the same
env so every tool this notebook calls resolves from `${syri_bin}`.

```{r install-tools, engine='bash'}
CONDA=/home/sr320/miniconda3

if [ ! -x "${syri_bin}/syri" ]; then
  "${CONDA}/bin/conda" create -y -n syri_env -c conda-forge -c bioconda \
    python=3.10 syri plotsr ragtag minimap2 samtools mummer4 bedtools numpy pandas
fi

"${syri}"      --version
"${ragtag}"    --version
"${minimap2}"  --version
"${dnadiff}"   -h 2>&1 | head -1
echo "tooling OK"
```

# STEP 1 — ANCHOR EACH ASSEMBLY TO THE REFERENCE CHROMOSOMES

RagTag `scaffold` aligns each ecotype's purged contigs to the reference and orders/orients them
into one scaffold per reference chromosome, named `<refchrom>_RagTag` (unplaced contigs keep their
original names). We then keep only the scaffolds corresponding to the **42 reference chromosomes**
(`NC_*` > 10 Mb) and strip the `_RagTag` suffix so their IDs match the reference exactly — SyRI
matches homologous sequences by identical name.

The contig→chromosome seed from [`21.1`](21.1-gene-anchored-synteny.Rmd)
(`output/21.1-gene-anchored-synteny/contig_to_chromosome.tsv`) already confirmed these contigs map
cleanly onto the 42 chromosomes; RagTag reproduces that placement at sequence level with
coordinates.

```{r anchor, engine='bash'}
cd "${output_dir}/anchor"

anchor_one () {   # $1 name   $2 purged.fa
  local name="$1" asm="$2"
  echo "=== RagTag scaffold: ${name} ==="

  "${ragtag}" scaffold "${ref_fasta}" "${asm}" \
    -o "${output_dir}/anchor/${name}" \
    -t "${threads}" \
    --aligner "${minimap2}" \
    -w \
    > "${output_dir}/logs/${name}.ragtag.log" 2>&1

  local scaf="${output_dir}/anchor/${name}/ragtag.scaffold.fasta"
  "${samtools}" faidx "${scaf}"

  # Keep only the 42 chromosome scaffolds (NC_*_RagTag); strip the _RagTag suffix so IDs
  # equal the reference chromosome names.
  grep -E '^NC_[0-9]+\.[0-9]+_RagTag' "${scaf}.fai" | cut -f1 \
    > "${output_dir}/anchor/${name}.placed.list"

  "${samtools}" faidx "${scaf}" $(cat "${output_dir}/anchor/${name}.placed.list") \
    | sed -E 's/^>(NC_[0-9]+\.[0-9]+)_RagTag/>\1/' \
    > "${output_dir}/anchor/${name}.anchored.fa"
  "${samtools}" faidx "${output_dir}/anchor/${name}.anchored.fa"

  awk -v n="${name}" '{s+=$2} END{print n" anchored: "NR" chromosomes, "s" bp"}' \
    "${output_dir}/anchor/${name}.anchored.fa.fai"
}

anchor_one lean     "${lean_fasta}"
anchor_one siscowet "${siscowet_fasta}"
```

Build the matching **42-chromosome reference subset** (drop the small `NC_*` mitochondrion and all
`NW_*` scaffolds) so the reference axis is the same 42 sequences the ecotypes were anchored to.

```{r ref-chroms, engine='bash'}
cd "${output_dir}/anchor"

"${samtools}" faidx "${ref_fasta}"
awk -v m="${min_chrom_bp}" '$1 ~ /^NC_/ && $2 > m {print $1}' "${ref_fasta}.fai" \
  > ref.chroms.list
wc -l ref.chroms.list   # expect 42

"${samtools}" faidx "${ref_fasta}" $(cat ref.chroms.list) > ref.chroms.fa
"${samtools}" faidx ref.chroms.fa
awk '{s+=$2} END{print "reference chromosomes: "NR", "s" bp"}' ref.chroms.fa.fai
```

# STEP 2 — WHOLE-GENOME ALIGNMENT (minimap2 asm5)

One `asm5` alignment per comparison, emitted as `--eqx` SAM (base-level `=`/`X` CIGAR, which SyRI
requires) and sorted to BAM. Target (`-r` for SyRI) is listed first, query (`-q`) second.

- `lean_vs_ref` : reference chromosomes ← lean
- `sisco_vs_ref`: reference chromosomes ← siscowet
- `lean_vs_sisco`: lean ← siscowet  (both share reference chromosome names after Step 1)

```{r wga, engine='bash'}
cd "${output_dir}/wga"

REF="${output_dir}/anchor/ref.chroms.fa"
LEAN="${output_dir}/anchor/lean.anchored.fa"
SISCO="${output_dir}/anchor/siscowet.anchored.fa"

align_pair () {   # $1 tag   $2 target.fa   $3 query.fa
  local tag="$1" t="$2" q="$3"
  echo "=== minimap2 asm5: ${tag} ==="
  "${minimap2}" -t "${threads}" -ax asm5 --eqx "${t}" "${q}" \
    2> "${output_dir}/logs/${tag}.minimap2.log" \
    | "${samtools}" sort -@ "${threads}" -O BAM -o "${output_dir}/wga/${tag}.bam" -
  "${samtools}" index "${output_dir}/wga/${tag}.bam"
  "${samtools}" flagstat "${output_dir}/wga/${tag}.bam" \
    > "${output_dir}/wga/${tag}.flagstat.txt"
  echo "  wrote ${tag}.bam"
}

align_pair lean_vs_ref   "${REF}"  "${LEAN}"
align_pair sisco_vs_ref  "${REF}"  "${SISCO}"
align_pair lean_vs_sisco "${LEAN}" "${SISCO}"
```

# STEP 3 — CALL STRUCTURAL VARIANTS (SyRI)

SyRI classifies every region of each alignment into **syntenic** blocks and **structural
rearrangements** (inversions `INV`, translocations `TRANS`/`INVTR`, duplications `DUP`/`INVDP`)
and, within aligned regions, **local variants** (SNPs, insertions `INS`, deletions `DEL`, copy
gains/losses `CPG`/`CPL`, highly-diverged regions `HDR`, tandem repeats `TDM`). Reference and
query IDs already match by construction (Step 1), so we run each comparison whole-genome; SyRI
processes homologous chromosomes and reports cross-chromosome placements as translocations.

```{r syri, engine='bash'}
cd "${output_dir}/syri"

REF="${output_dir}/anchor/ref.chroms.fa"
LEAN="${output_dir}/anchor/lean.anchored.fa"
SISCO="${output_dir}/anchor/siscowet.anchored.fa"

call_svs () {   # $1 tag   $2 target.fa   $3 query.fa
  local tag="$1" t="$2" q="$3"
  echo "=== SyRI: ${tag} ==="
  "${syri}" \
    -c "${output_dir}/wga/${tag}.bam" \
    -r "${t}" \
    -q "${q}" \
    -F B \
    --nc "${threads}" \
    --dir "${output_dir}/syri" \
    --prefix "${tag}." \
    > "${output_dir}/logs/${tag}.syri.log" 2>&1

  echo "  ${tag}.syri.out rows: $(wc -l < ${output_dir}/syri/${tag}.syri.out)"
  cut -f11 "${output_dir}/syri/${tag}.syri.out" | sort | uniq -c | sort -rn
}

call_svs lean_vs_ref   "${REF}"  "${LEAN}"
call_svs sisco_vs_ref  "${REF}"  "${SISCO}"
call_svs lean_vs_sisco "${LEAN}" "${SISCO}"
```

# STEP 4 — MERGED SV TABLE & SUMMARY

Parse the three `*.syri.out` files into one tidy, typed SV table plus a per-comparison,
per-type summary. `*.syri.out` is 11-column TSV (no header): reference `chr/start/end`, ref &
query sequence, query `chr/start/end`, unique id, parent id, and the **annotation type** (col 11).
We keep rows with numeric reference coordinates and tag each with a coarse class.

```{r merged-sv-table, eval=TRUE}
library(tidyverse)

syri_dir <- file.path(output_dir, "syri")

# Coarse classes over SyRI's fine-grained annotation types.
sv_class <- tribble(
  ~type,   ~class,
  "SYN",   "syntenic",
  "SYNAL", "syntenic",
  "INV",   "inversion",     "INVAL",  "inversion",
  "TRANS", "translocation", "TRANSAL","translocation",
  "INVTR", "translocation", "INVTRAL","translocation",
  "DUP",   "duplication",   "INVDP",  "duplication",
  "INS",   "indel",         "DEL",    "indel",
  "CPG",   "copy_var",      "CPL",    "copy_var",
  "HDR",   "diverged",      "TDM",    "tandem_repeat",
  "SNP",   "snp",
  "NOTAL", "unaligned"
)

read_syri <- function(tag) {
  path <- file.path(syri_dir, paste0(tag, ".syri.out"))
  if (!file.exists(path)) return(tibble())
  read_tsv(
    path,
    col_names = c("ref_chr", "ref_start", "ref_end", "ref_seq", "qry_seq",
                  "qry_chr", "qry_start", "qry_end", "uid", "parent", "type"),
    col_types = cols(.default = "c"),
    na = c("", "-", "NA"),
    show_col_types = FALSE
  ) %>%
    mutate(comparison = tag,
           ref_start = suppressWarnings(as.numeric(ref_start)),
           ref_end   = suppressWarnings(as.numeric(ref_end)),
           qry_start = suppressWarnings(as.numeric(qry_start)),
           qry_end   = suppressWarnings(as.numeric(qry_end)))
}

sv <- c("lean_vs_ref", "sisco_vs_ref", "lean_vs_sisco") %>%
  map_dfr(read_syri) %>%
  left_join(sv_class, by = "type") %>%
  mutate(class = replace_na(class, "other"))

# Full merged table (all rows, coarse-classed)
readr::write_tsv(sv, file.path(output_dir, "syri", "merged_sv_table.tsv"))

# Per-comparison / per-class summary (structural + local variants only; drop syntenic backbone)
sv_summary <- sv %>%
  filter(!class %in% c("syntenic", "unaligned"), !is.na(ref_start)) %>%
  mutate(ref_len = pmax(ref_end - ref_start, 0)) %>%
  group_by(comparison, class) %>%
  summarise(n = n(), total_ref_bp = sum(ref_len, na.rm = TRUE), .groups = "drop") %>%
  arrange(comparison, desc(n))

readr::write_tsv(sv_summary, file.path(output_dir, "syri", "sv_summary.tsv"))
print(sv_summary, n = Inf)
```

# STEP 5 — PLOTS

Two figure families, written to `analyses/23-genome-sv/plots/`:

1. **SV-class summary bars** from `sv_summary.tsv` (counts and bp per comparison).
2. **Genome-wide structural-rearrangement plots** with `plotsr`, one per comparison, showing
   syntenic / inverted / translocated / duplicated segments across the 42 chromosomes.

```{r plots-summary, eval=TRUE}
library(tidyverse)

plot_dir <- file.path(output_dir, "plots")
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

sv_summary <- read_tsv(file.path(output_dir, "syri", "sv_summary.tsv"),
                       show_col_types = FALSE)

comp_levels <- c("lean_vs_ref", "sisco_vs_ref", "lean_vs_sisco")

p_counts <- sv_summary %>%
  mutate(comparison = factor(comparison, levels = comp_levels)) %>%
  ggplot(aes(x = class, y = n, fill = comparison)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  scale_y_log10() +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "Structural-variant counts by class",
       x = NULL, y = "SV count (log10)", fill = "Comparison")

ggsave(file.path(plot_dir, "01_sv_counts.png"), p_counts, width = 10, height = 5, dpi = 300)
ggsave(file.path(plot_dir, "01_sv_counts.svg"), p_counts, width = 10, height = 5)

p_bp <- sv_summary %>%
  mutate(comparison = factor(comparison, levels = comp_levels)) %>%
  ggplot(aes(x = class, y = total_ref_bp / 1e6, fill = comparison)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "Structural-variant span on reference coordinates",
       x = NULL, y = "Total reference span (Mb)", fill = "Comparison")

ggsave(file.path(plot_dir, "02_sv_span_mb.png"), p_bp, width = 10, height = 5, dpi = 300)
ggsave(file.path(plot_dir, "02_sv_span_mb.svg"), p_bp, width = 10, height = 5)

list.files(plot_dir)
```

```{r plots-plotsr, engine='bash'}
cd "${output_dir}/plots"

REF="${output_dir}/anchor/ref.chroms.fa"
LEAN="${output_dir}/anchor/lean.anchored.fa"
SISCO="${output_dir}/anchor/siscowet.anchored.fa"

# plotsr reads a genomes.txt (fasta<TAB>display-name, in plotting order) plus one --sr per
# adjacent pair. We draw each pairwise comparison on its own.
plotsr_pair () {   # $1 tag   $2 target.fa  $3 target-name   $4 query.fa  $5 query-name
  local tag="$1"
  printf '%s\t%s\n%s\t%s\n' "$2" "$3" "$4" "$5" > "${tag}.genomes.txt"
  "${plotsr}" \
    --sr "${output_dir}/syri/${tag}.syri.out" \
    --genomes "${tag}.genomes.txt" \
    -H 8 -W 10 \
    -o "${output_dir}/plots/03_plotsr_${tag}.png" \
    > "${output_dir}/logs/${tag}.plotsr.log" 2>&1 \
    && echo "  wrote 03_plotsr_${tag}.png" \
    || echo "  plotsr ${tag} failed (see logs/${tag}.plotsr.log)"
}

plotsr_pair lean_vs_ref   "${REF}"  ref  "${LEAN}"  lean
plotsr_pair sisco_vs_ref  "${REF}"  ref  "${SISCO}" siscowet
plotsr_pair lean_vs_sisco "${LEAN}" lean "${SISCO}" siscowet
```

# STEP 6 — (OPTIONAL) dnadiff IDENTITY REPORT

The plan calls for a `dnadiff` alignment-identity/coverage report. This uses the MUMmer4 `nucmer`
in `syri_env` (threaded, unlike the v3.1 at `/usr/bin`). It is the heaviest step here and is not
required for SV calling or the PAV cross-validation — run it for the manuscript's assembly-identity
numbers. The `.report` files carry AvgIdentity, aligned bases, and rearrangement/indel/SNP counts.

```{r dnadiff, engine='bash'}
cd "${output_dir}/wga"

REF="${output_dir}/anchor/ref.chroms.fa"
LEAN="${output_dir}/anchor/lean.anchored.fa"
SISCO="${output_dir}/anchor/siscowet.anchored.fa"

run_dnadiff () {   # $1 tag  $2 ref.fa  $3 qry.fa
  local tag="$1"
  echo "=== dnadiff: ${tag} ==="
  "${dnadiff}" -p "${output_dir}/wga/${tag}.dnadiff" "$2" "$3" \
    > "${output_dir}/logs/${tag}.dnadiff.log" 2>&1
  grep -E 'AvgIdentity|AlignedBases|TotalIndels|TotalSNPs' \
    "${output_dir}/wga/${tag}.dnadiff.report" || true
}

run_dnadiff lean_vs_ref   "${REF}"  "${LEAN}"
run_dnadiff sisco_vs_ref  "${REF}"  "${SISCO}"
run_dnadiff lean_vs_sisco "${LEAN}" "${SISCO}"
```

# CROSS-VALIDATION WITH REFERENCE-BASED PAV

Once the reference comparisons (`lean_vs_ref`, `sisco_vs_ref`) have SyRI output, run
[`code/23.2-sv-pav-crossvalidate.py`](23.2-sv-pav-crossvalidate.py). It intersects the
reference-based PAV calls from [`15`](15-diff-pav.py) with SyRI-confirmed SV regions (both on
reference coordinates) and emits the high-confidence set.

```{r crossvalidate, engine='bash'}
cd "${project_dir}"
python3 code/23.2-sv-pav-crossvalidate.py
```

# CHECKSUMS

```{r checksums, engine='bash'}
cd "${output_dir}"
for f in anchor/lean.anchored.fa anchor/siscowet.anchored.fa anchor/ref.chroms.fa \
         syri/lean_vs_ref.syri.out syri/sisco_vs_ref.syri.out syri/lean_vs_sisco.syri.out \
         syri/merged_sv_table.tsv syri/sv_summary.tsv; do
  [ -s "$f" ] && md5sum "$f" | tee "$f".md5
done
```

# DELIVERABLES

Written to `analyses/23-genome-sv/`:

| File | Description |
|---|---|
| `anchor/{lean,siscowet}.anchored.fa` | Purged assembly RagTag-scaffolded onto the 42 reference chromosomes (IDs = reference chromosome names) |
| `anchor/ref.chroms.fa` | 42-chromosome reference subset (shared axis) |
| `wga/{tag}.bam` | minimap2 `asm5` whole-genome alignment per comparison |
| `syri/{tag}.syri.out` / `.vcf` | SyRI structural + local variant calls per comparison |
| `syri/merged_sv_table.tsv` | All SVs, coarse-classed, across the three comparisons |
| `syri/sv_summary.tsv` | Per-comparison / per-class SV counts and reference span |
| `plots/01_sv_counts.*`, `02_sv_span_mb.*` | SV summary bars |
| `plots/03_plotsr_{tag}.png` | Genome-wide structural-rearrangement plots |
| `wga/{tag}.dnadiff.report` | (Optional) alignment identity/coverage report |

These feed the reference-vs-SyRI PAV cross-validation
([`23.2`](23.2-sv-pav-crossvalidate.py)) and the browser/manuscript integration in Phase 5 of
[`code/23-next-phase-research-plan.md`](23-next-phase-research-plan.md).

# REFERENCES
